---
title: Ecological perspectives on technological diversity at Kanjera South
author:
  - name: Jonathan S. Reeves
    email: jonathan.reeves@uni-tuebingen.de
    affiliation: Department of Prehistory and Quarternary Ecology, Eberhard Karls University of Tübingen
    footnote: 1
  - name: David R. Braun
    email: David_Braun@gwu.edu
    affiliation: Center for the Advanced Study of Human Paleobiology, The George Washington University
    footnote: 2
  - name: Emma M. Finestone
    email: finestone@shh.mpg.de
    affiliation: Department of Archaeology, Max Planck Institute for the Science of Human History
    footnote: 3
  - name: Thomas W. Plummer
    email: thomas.plummer@qc.cuny.edu
    affiliation: Department of Anthropology, Queens College
    footnote: 4
address:
  - code: Department of Prehistory and Quarternary Ecology
    address: Eberhard Karls University of Tübingen, Schloss Hohentübingen Burgsteige 11, 72070 Tübingen, Germany
  - code: Center for the Advanced Study of Human Paleobiology
    address: George Washington University, 800 22nd Street, North West, Washington D.C., USA.
  - code: Department of Archaeology
    address: Max Planck Institute for the Science of Human History, Kahlaische Strasse 10, D-07743, Jena, Germany
  - code: Department of Anthropology, Queens College
    address: CUNY & NYCEP, 65-30 Kissena Blvd, Flushing, NY 11367, USA
footnote:
    
  - code: 5
    text: "Corresponding Author"
abstract: |
  The aspects of hominin behavior responsible for Oldowan stone tool variation has been the focus of much debate. There is some consensus that Oldowan artifact variation arises from a combination of ecological and cultural factors. These factors are often examined independently of one another. The diversity of raw material types and technological strategies present at the site of Kanjera South provide an opportunity to examine the interacting effect of ecology and culture on Oldowan stone tool variation. Here we combine previous analyses of raw material properties, provenance, and technology with quantitative measures of core reduction intensity and tool utilization to examine the influence of both ecological and techno-cultural factors on stone tool variation at Kanjera South. The results of this analysis show that technological variation at a Kanjera South reflects a dynamic relationship between raw material properties, provenance, and hominin mobility. Cores produced on raw materials from distant sources are more reduced than locally sourced raw materials. Distant raw materials are generally more resistant to edge attrition compared to those available locally which may have incentivized their transport over long distances. Moreover, the variation in stone tool reduction is not constistent with neutral models of stone tool transport and discard. This suggests that the lithic assemblage at Kanjera South may reflect a structured land-use strategy that may relate to the resource rich nature of the Homa Penninsula. This pattern of stone tool utilization also has an impact on the technological strategies employed by Oldowan tool makers at Kanjera South. Cores produced on raw materials from distant sources also exhibit more complex core reduction strategies than locally acquired materials. While this pattern is partially due to the differences the quality of knappable stone, bifacial centripetal and multifacial core reduction strategies also arise due to the continuous transport and use of exotic raw materials. These results demonstrate that ecological factors such as raw material provenance and physical properties have strong impacts on reduction intensity and the technological strategies utilized by hominins. Yet, not all stone tool variation at Kanjera South can be explained by this relationship. These results suggest that Oldowan stone tool variation should not be examined from a strictly ecological or technological perspective, but rather within the context of its broader cultural-ecological system.
journal: "Journal of Human Evolution"
date: "`r Sys.Date()`"
bibliography: Kanjera_lithics_JHE.bib
#linenumbers: true
#numbersections: true
header-includes: 
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
output: 
   word_document: 
      reference_docx: Manuscript_Template.docx
   pdf_document: rticles::elsevier_article
---

\doublespacing
\pagebreak
\linenumbers

```{r step up, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

tiff_figures = TRUE

library(ggplot2)
library(grid)
library(gridExtra)
library(dplyr)
library(plyr)
library(citr)
library(FSA)
library(dunn.test)
library(knitr)
library(knitcitations)

cores <- read.csv("Data/CORES_2019_JHE_With_REDUX")
cores$RM <- toupper(cores$RM)
```

\doublespacing

# Introduction


  The Oldowan is relatively simplistic as it consists primarily of core and flake tools [@leakeyOlduvaiGorgeVol1971]. In contrast to earlier typological approaches, experimental models by [@tothOldowanReassessedClose1985] suggest that the Oldowan is a simple core and flake industry whose primary goal is the removal of flakes with sharp cutting edges. In spite of its superficial simplicity, Oldowan stone tool variation is a vast resource of information regarding Early Pleistocene hominin behavior. Technological analyses show that Oldowan hominins had at least a basic understanding of the general principles of flaking and selection of suitable tool stones for artifact manufacture [@braunEarliestKnownOldowan2019; @braunRawMaterialQuality2009; @delagnesLatePlioceneHominid2005; @delatorreOmoRevisitedEvaluating2004; @rocheEarlyHominidStone1999; @semawWorldOldestStone2000; @stoutRawMaterialSelectivity2005]. In addition, Oldowan hominins also transported stone tools various distance to the places where they are used and discarded [@blumenschineArchaeologicalPredictionsHominid1998; @braunOldowanBehaviorRaw2008; @harmandVariabilityRawMaterial2009; @hayGeologyOlduvaiGorge1976; @isaacArchaeologyHumanOrigins1984; @pottsWhyOldowanPlioPleistocene1991; @tothOldowanReassessedClose1985]. Despite early suggestions that the Oldowan was a largely expedient tool kit [@chavaillonDecouverteNiveauOldowayen1970], it may reflect a more nuanced technical system, where raw material is acquired, transported, utilized, maintained and eventually discarded [@isaacArchaeologyHumanOrigins1984].
        
A major difficulty in the study of Oldowan hominin behavior is linking variation in artifacts to individual influences on behavior [@gallottiAcheuleanEastAfrica2018; @rocheOriginsAdaptationsEarly2009]. While socio-cognitive approaches only require examining technological strategies present in the artifacts themselves, ecological analyses require integrated ecological and functional data sets. Demonstrating the influence of ecological parameters on stone tool use requires establishing spatial relationships between measures of stone tool utilization and landscape features such as raw material sources and paleogeographic settings [@blumenschineArchaeologicalPredictionsHominid1998; @blumenschineEffectsDistanceStone2008; @blumenschineOlduvaiGorgeOlduvai2012; @braunLandscapescaleVariationHominin2008; @braunOldowanTechnologyRaw2009; @isaacKoobiForaResearch1997; @rogersLandscapeArchaeologicalInvestigation1997]. While previous work on this topic has demonstrated the influence of raw material access on Oldowan assemblages [@blumenschineEffectsDistanceStone2008], little work has been done demonstrating its direct influence on technological variation in the Oldowan. The site of Kanjera South provides an opportunity to do so. Extensive research on the technological context has characterized not only the technological strategies employed by hominins, at this site, but also the landscape context of the site [@behrensmeyerPleistoceneLocalityKanjera1995; @braunOldowanBehaviorRaw2008; @braunOldowanTechnologyRaw2009; @braunRawMaterialQuality2009; @ditchfieldGeochronologyPhysicalContext2018; @bishopRecentResearchOldowan2006; @kentPleistoceneBedsKanam1942a; @lemoriniOldStonesSong2014; @lemoriniOldStonesSong2019; @oliverBovidMortalityPatterns2019; @plummerEnvironmentalContextOldowan2009; @plummerLatePlioceneOldowan2001; @plummerOldowanHomininBehavior2016; @plummerResearchLatePliocene1999]. Here we present an integrated approach that examines Oldowan lithic technology from an ecological context. The results of this study contribute to the exploration of ecological and socio-cultural influences on Oldowan artifact variation.

# Background
 
Given the dynamic system that Oldowan tools ultimately reflect, it is argued that stone tool variation is influenced by function, ecology, culture and cognition [@gallottiAcheuleanEastAfrica2018; @isaacArchaeologyHumanOrigins1984; @plummerEnvironmentalContextOldowan2009; @tothOldowanCaseStudies2006] (Figure \ref{variation}). Understanding the role of function in shaping Oldowan stone tool variation is constrained. The uses of Oldowan tools beyond evidence of cutting, and flake production is not well known. A small number of use-wear studies are beginning to shed light on the diversity of functions that Oldowan flakes were used for [@keeleyMicrowearPolishesEarly1981; @lemoriniOldStonesSong2014; @lemoriniOldStonesSong2019]. The impact of these functions on Oldowan stone tool variation remains unclear. As a result, Oldowan studies have focused primarily on drawing behavioral inferences from explorations of hominin ecology, culture and cognition derived from the morphological variation in stone tools. However, these various aspects of hominin life-ways are seldom studied in tandem. 

```{r fig-1, echo=FALSE, warning=FALSE, , fig.cap="A diagram of factors that contribute to stone tool variation. Adopted from Isaac (1984). \\label{variation}", paged.print=FALSE}
library(png)
library(grid)

  figure_03 = readPNG("Figures/Lithics_Form_Figure_2.png")
  g <- rasterGrob(figure_03, interpolate=TRUE)
  ggplot(NULL, aes(NULL, NULL)) + geom_blank() + xlab('') + ylab('') +
    annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
    theme(panel.background = element_blank())

```

The ecological approach views tools as an extra-somatic mechanism for solving environmental problems [@binfordConstructingFramesReference2001]. This approach was originally championed by the work of Glynn Isaac, John Desmond Clark, and their students [@clarkAfricaPrehistoryPeripheral1975; @harrisKarariIndustryIts1978; @isaacKoobiForaResearch1997; @shickModelingFormationEarly1987; @tothBehavioralInferencesEarly1987; @tothStoneTechnologiesEarly1982]. In this perspective technology is considered to be a dynamic process that is adapted to the temporal and spatial distribution of resources [@bamforthSettlementRawMaterial1990; @isaacScatterPatches1976; @isaacStoneAgeVisiting1981; @ludwigTechnologicalReassessmentEast1998; @nelsonStudyTechnologicalOrganization1991; @petersLandscapePerspectivesPossible1995; @rogersChangingPatternsLand1994; @tothOldowanReassessedClose1985]. In essence this approach investigates artifacts as dynamic objects that are a part of the hominin foraging system. One component of this approach is the concept that artifacts reflect movement of stone resources across landscapes often referred to as the “the flow of stone” model [@isaacArchaeologyHumanOrigins1984]. Some artifacts are produced and discarded whereas others are transported away from the site. Understanding how this flow of stone through structured landscapes produces inter-site assemblage variability provides insight into the ecology and foraging strategies of early hominins [@blumenschineOlduvaiGorgeOlduvai2012]. As a result, a variety of models have been proposed and debated for how assemblage composition, lithic attributes and artifact density reflect land-use patterns at multiple spatial scales [@binfordContemporaryModelBuilding1976; @foleyModelRegionalArchaeological1981; @foleyOffsiteArchaeologyAlternative1981; @isaacBonesContentionCompeting1983; @isaacScatterPatches1976; @isaacStoneAgeVisiting1981; @shickModelingFormationEarly1987; @tothBehavioralInferencesEarly1987]. Ecologically oriented research often measures aspects of stone tool variation in relation to measures of tool utilization and transport. Quantitative or ordinal variables such as mass, length, proportion of cortex, and flake scar density are often compared alongside landscape variables such as distance to raw material source and paleogeographic context [@andrefskyAnalysisStoneTool2009; @andrefskyRawMaterialAvailabilityOrganization2008; @blumenschineOlduvaiGorgeOlduvai2012; @bunnFxJj50EarlyPleistocene1980; @daviesModelingRelationshipsSpace2018; @douglassArchaeologicalPotentialInformal2010; @holdawayRecordStoneStudy2004; @brantinghamMobilityCompetitionPlioPleistocene1998; @kuhnWhatMovesUs2016; @tothBehavioralInferencesEarly1987]. Examples of this type of approach includes studies of artifact variation throughout the Okote Member of the Koobi Fora Formation. Paleogeographic context appears to influence the transport patterns and reduction intensity of these artifacts when they are discarded [@braunLandscapescaleVariationHominin2008; @rogersChangingPatternsLand1994; @rogersLandscapeArchaeologicalInvestigation1997; @sternFxJj43Window5Million2002]. At Olduvai, quartz artifacts vary in size and abundance according to their distance from a large outcrop of quartz called Naibor Soit [@blumenschineEffectsDistanceStone2008].Measures of cortex and core reduction have shed light on the dynamic input-output system under which Oldowan lithic assemblages form [@bunnFxJj50EarlyPleistocene1980; @rogersLandscapeArchaeologicalInvestigation1997; @tothBehavioralInferencesEarly1987; @braunLandscapescaleVariationHominin2008].

   An alternative approach to the ecological focus, is one where the actions and thoughts associated with stone tool production are highlighted [@selletChaineOperatoireConcept1993]. This is sometimes referred to as the technological approach [@soressi2011reduction]. From this perspective, the variation seen in all technology stems from socially mediated images and thoughts [@inizanTechnologyTerminologyKnapped1999]. The various combinations of ways to flake, shape or retouch stone are argued to reflect the different knapping strategies that characterize the skills of a group or individual [@delagnesLatePlioceneHominid2005]. From this perspective stone tools reflect the cognitive capacity and the technical competence of Oldowan tool makers [@rocheOriginsAdaptationsEarly2009; @schick1994making; @stoutCognitiveDemandsLower2015; @stoutMakingToolsMaking2009; @tothOverviewCognitiveImplications2018]. In particular, certain components of the artifact production process (e.g. complex series of nested goals) appear to require specific cognitive traits [@wynnArcheologicalInsightsHominin2016]. Oldowan variation is interpreted as differences in skills, cultural traditions, or even different taxa [@delagnesLatePlioceneHominid2005; @rocheEarlyHominidStone1999; @rocheNaiyenaEngolWest2018; @rocheOriginsAdaptationsEarly2009; @stoutTechnologicalVariationEarliest2010]. Technological analyses have been instrumental in revealing the rules that govern stone artifacts production. Analysis of lithic assemblages from West Turkana in Kenya show that hominins had already mastered flake removal and how to maintain and exploit pre-existing platforms [@delagnesLatePlioceneHominid2005].

Moreover, differences in the arrangement of flake scars, refit sequences, the frequency of knapping accidents are used to infer different production strategies from Oldowan lithic assemblages. Differences in the inferred skill of toolmakers at Lokalelei 1 and 2c are so striking that the author suggested that the differences arise from 2 different social groups or possibly different species of hominins [@delagnesLatePlioceneHominid2005]. At Gona, differences in the proportion of reduction strategies between sites have been argued to reflect cultural traditions of specific groups [@stoutTechnologicalVariationEarliest2010; @stoutArchaeologyOriginsHuman2019]. At a broader scale, some researchers have argued that the variation in the Oldowan is similar to the regional variation in the observed in chimpanzee cultures [@barhamFirstAfricansAfrican2008; @whitenCulturesChimpanzees1999; @whitenEvolutionCulturalTransmission2009a]. The social context within this variation is interpreted to reflect potentially different social learning mechanisms in the archaeological record [@morganExperimentalEvidenceCoevolution2015; @stoutArchaeologyOriginsHuman2019].
 
The vast majority of research on Oldowan artifacts can be identified as falling into one the previously described categories (i.e. either ecologically focused or socially mediated). The singular focus of most Oldowan research on either ecological or technological questions has an impact on the current status of Oldowan research. The manner in which the Oldowan is described, and the methods used to describe it, depend heavily on the theoretical position of the analyst [@detorreRemarksCurrentTheoretical2009]. In a way, this is problematic because as figure \ref{variation} demonstrates, the factors that contribute to stone tool variability are not mutually exclusive and likely interact with one another For example, the strategy used to reduce a core will affect the proportion of different Toth types produced [@stoutArchaeologyOriginsHuman2019; @stoutTechnologicalVariationEarliest2010]. This will also likely influence how cortex is distributed across a landscape [@tothBehavioralInferencesEarly1987]. Therefore in the absence of a more integrated approach, differences in the representation of flake types (i.e. Toth Types) between two sites could be interpreted as varied foraging techniques when in reality they result from different tool reduction strategies (or possibly a combination of both). Therefore, in this example, reduction strategy and measurements of cortex must both be considered before discussing the aspects of behavior reflected in a given assemblage. In other words, without considering the broader ecological context of the Oldowan it may be impossible to disentangle technical competence of Oldowan knapppers from technological variability (or _vice versa_). 

Distance - decay relationships between the location of tool-stone sources and stone tool discard is prevalent in Oldowan assemblages [@blumenschineEffectsDistanceStone2008; @braunLandscapescaleVariationHominin2008; @tothOldowanReassessedClose1985]. However, these patterns are often described in terms of reduction intensity, size of flakes and cores, or the representation of cortex. Oldowan cores that are more reduced often involve greater levels of core rotation [@tothStoneTechnologiesEarly1982]. It is possible that differences in technological strategies may sometimes arise due to raw material abundance. Moreover, it may be that Oldowan stone tool variation may be the result of a complex interaction between social mechanisms of information transfer and ecological context. Understanding the significance of Oldowan variability requires an integrated approach which considers the context of tool use as well as the context of availability of resources. The assemblage of Oldowan artifacts from Kanjera South provide the opportunity to understand an artifact record within the context of raw material availability and technological variability [@plummerOldowanHomininBehavior2016].

# Background to Kanjera South 

```{r fig-2, fig.height=5, fig.width=7, fig.cap= "A map of the Homa Peninsula. Kanjera South is situated to the East of Homa Mountain. The Homa Mountain carbonatite center is the primary source of the local raw materials including Homa limestone (HLi), Homa Phonolite (HPh), and Fenetized nyanzian rocks (FNy). Distant or exotic raw materials originate much farther to the east but they can be found in drainages east of the Samanga Fault. These include Bukoban andesite (BBa), Bukoban felsite (BFe), Bukoban quartzite (BQu), Nyanzian rhyolite (NyR), and Oygus granite (OGr)  \\label{map}" }

library(raster)
library(ggplot2)
library(ggsn)
library(ggsflabel)
library(ggrepel)
library(sf)
library(ggnewscale)

dem <- raster("Figures/Figure_data/Homa_Penninsula_DEM_30m2.tif")

dem[dem$Homa_Penninsula_DEM_30m2 <= 1134] <- NA
slope <- raster::terrain(dem, opt = "slope")
aspect <- terrain(dem, opt = "aspect")
dem <- data.frame(rasterToPoints(dem))
hs <- hillShade(slope = slope,
                aspect = aspect,
                angle = 50,
                direction = 315,
                normalize = TRUE)

hs <- data.frame(rasterToPoints(hs))

water <- read_sf("Figures/Figure_data/ke_major-rivers.shp")
kjs <- read_sf("Figures/Figure_data/KJS_WSG_84.shp")
rivers2<-read_sf("Figures/Figure_data/Rivers_homa_WGS2.shp")
fault <- read_sf("Figures/Figure_data/SamangaFault_WGS.shp")
st_geometry(kjs) <- NULL

ggplot()+
   geom_raster(data = hs,aes(x=x, y=y, fill = layer),show.legend = FALSE)+
   
   scale_fill_gradientn(colours = grey.colors(100,start = .5, end = 0))+
   
   new_scale_fill()+
   
   geom_raster(data = dem[dem$Homa_Penninsula_DEM_30m2 > 1134 & dem$Homa_Penninsula_DEM_30m2 < 2000,],
               aes(x=x,y=y,fill = Homa_Penninsula_DEM_30m2),alpha = .4, inherit.aes = FALSE)+
  
  #geom_sf(data = water, color = "dodger blue1")+
  
  geom_sf(data = rivers2, color = "dodger blue2")+
  
  geom_sf(data = fault,linetype =2) +
  geom_sf_text(data = fault, aes(label = "Samanga Fault"), angle = 47, nudge_y = -.015)+ 
      
   scale_fill_gradientn(colours = terrain.colors(100), "Elevation (m)")+
  
    geom_point(data = kjs, aes(x= X, y = Y)) +
   
   geom_label_repel(data= kjs, aes(x = X, y = Y,label = Local),
                  box.padding   = 1, 
                  point.padding = .5,
                  segment.color = 'black',
                  color = "black")+
   
   north(location = "bottomleft", symbol = 10, x.min = 34.43,x.max = 34.9, y.min = -0.525, y.max = -.27195)+
   
   scalebar(location = "bottomleft",dist_unit = "km",dist = 4, x.min = 34.44,x.max = 34.9, y.min = -0.528, y.max = -.275, transform = TRUE,st.dist = .025) + 
   
   scale_x_continuous(limits = c(34.42,34.8), expand = c(0,0))+
   
   scale_y_continuous(limits = c(-0.55,-.275), expand = c(0,0))+
   
   theme(panel.background = element_rect(fill = "dodger blue1"),
         panel.grid = element_blank(),
         panel.spacing = unit(1, "cm"),
         axis.text = element_blank(),
         axis.title = element_blank(),
         axis.ticks = element_blank())


```
   
   The site of Kanjera South is situated on the northern side of the Homa Peninsula on the edges of the Nyanza Rift near the shores of Lake Victoria [@plummerOldestEvidenceToolmaking2009; @plummerResearchLatePliocene1999]. The archaeological site is in an amphitheater of sediment that has been exposed by the Kanjera River that is draining off the Homa Mountain carbonatite complex (figure: \\ref{map}). This volcanic edifice dominates the sedimentary processes and basement geology of the region (LeBas 1970). The site of Kanjera South was excavated from a 3 meter stacked sequence of clays and silts that record extensive hominin behavior in the form of stone artifacts and bones with surface modifications that are indicative of the butchery of meat and acquisition of marrow [@braunOldowanTechnologyRaw2009; @plummerOldestEvidenceToolmaking2009]. Extensive excavation has recovered over 3000 fossils and similar numbers of stone artifacts [@plummerEnvironmentalContextOldowan2009]. The stratigraphy at Kanjera South is made up of approximately 30 m of fluvial, colluvial and lacustrine sediments [@ditchfieldGeochronologyPhysicalContext2018]. This stratigraphic sequence is well dated to ~2. Ma using a combination of paleomagnetic and biochronological markers.  These sediments are formally assigned to the Southern Member of the Kanjera Fm [@behrensmeyerPleistoceneLocalityKanjera1995; @plummerResearchLatePliocene1999]. There are five beds at Kanjera South (from oldest to youngest KS-1 to KS-5) with artifacts and fossils concentrated in the sands and silts of upper KS-1 to KS-3 [@ditchfieldGeochronologyPhysicalContext2018].  Lithics and fossils are thought to have been predominantly accumulated by hominins except for those found within thin, discontinuous conglomerates which represent brief intervals of higher energy water flow [@ditchfieldGeochronologyPhysicalContext2018; @ferraroEarliestArchaeologicalEvidence2013].

   The lithic assemblage recovered from exacavations at Kanjera South represents one of the largest single accumulations of Oldowan artifacts in association with modified fossil bone [@plummerFlakedStonesOld2004]. The largest site (169 m^2^) yielding the bulk of the archaeological finds was Excavation 1. The frequencies of different bovids in these beds indicates that the Kanjera South landscape, unlike the setting of most of the Oldowan sites, was dominated by a grasslands as opposed to trees [@plummerOldestEvidenceToolmaking2009].  This is further supported by the stable isotope analysis of pedogenic carbonates and tooth enamel indicating a relatively high proportion of plants that use a C4 photosynthetic pathway. The fossils associated with the stone artifacts at Kanjera South exhibit substantial evidence of hominin involvement in the accumulation of materials [@ferraroEarliestArchaeologicalEvidence2013]. Evidence from bone surface modifications as well as the proportions of bones modified are similar to patterns seen in "hominin first" experimental datasets for small mammals, while larger mammals show more mixed access [@ferraroEarliestArchaeologicalEvidence2013]. The totality of the zooarchaeological evidence at Kanjera strongly implicates a scenario where hominins had early access to small carcasses. Evidence from mortality profile data also supports the inference that hominins may have been able to access smaller carcasses prior to carnivores and that larger carcasses were at times scavenged [@oliverBovidMortalityPatterns2019].
  
```{r fig-3, echo=FALSE, warning=FALSE, , fig.cap="Examples of the found at Kanjera South. (A) Core produced on Homa limestone. (B) Core produced on Oyugis granite. (C) Core produced on Bukoban quartzite. (D) Core produced on Fenetized nyanzian.  \\label{tools}", paged.print=FALSE}
library(png)
library(grid)

  figure_03 = readPNG("Figures/KJS_Lithics_Fig_PNG.png")
  g <- rasterGrob(figure_03, interpolate=TRUE)
  ggplot(NULL, aes(NULL, NULL)) + geom_blank() + xlab('') + ylab('') +
    annotation_custom(g, xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf)+
    theme(panel.background = element_blank())
```
  
   The stone artifacts from Kanjera South reflect a wide diversity of technological adaptations [@braunOldowanTechnologyRaw2009]. The high diversity of the local geology on the Homa Peninsula provided a wide choice of different raw material types that hominins could incorporate into their toolkit . There are 16 formally described raw material types with many additional types that only appear in small quantities within the assemblage [@braunOldowanTechnologyRaw2009]. A detailed provenance study of the rock types used to make stone artifacts at Kanjera documented two features that distinguish the Kanjera South assemblage from many other Oldowan assemblages. The first is the presence of transport to Kanjera South from conglomerates that are at least 10 km away from the site [@braunOldowanBehaviorRaw2008]. The second is the systematic selection of rocks that have specific mechanical properties [@braunRawMaterialQuality2009]. Hominins selected rocks that fracture consistently and also had edges that resisted wear. The selection of certain rock types exceeds that which is seen in other Oldowan assemblages. Furthermore the intensity of selective behaviors seems to parallel rock properties. These differences in rock properties are reflected in the flaking technology (figure \ref{tools}). Unidirectional and multidirectional flaking patterns feature prominently in this assemblage. However, the flaking patterns track raw materials differences and there appears to be a concerted effort to extend the use life of cores in some raw materials by increasing the length of flaking sequences in some raw materials. These different technological strategies applied to different raw materials results in a high diversity of core reduction strategies in the Kanjera South Assemblage [@braunOldowanTechnologyRaw2009]. 
   
   In addition to information regarding the transport and selection of certain raw materials, detailed studies of the micro-wear on the stone artifacts from Kanjera provide insights into artifact use [@lemoriniOldStonesSong2014; @lemoriniOldStonesSong2019]. Usewear evidence suggests that the tool used by hominin at Kanjera were used for a variety of different activities. Most of these actions involve cutting, and the different raw materials appear to be used on different substrates. Usewear studies also suggests that some of the stone tools were used for modifying wood. This suggests that hominins were using stone artifacts to fashion other wooden tools [@lemoriniOldStonesSong2019]. Combined with the transport information, the sum of the information about stone artifacts at Kanjera South suggests that stone artifacts represented a significant component of the extractive foraging adaptation of Oldowan hominins. 

# Methods
                
   Oldowan tools undergo a series of transformations, through the removal of flakes, from the time they are acquired until they are ultimately discarded [@andrefskyAnalysisStoneTool2009]. As reviewed above, the flake removal process relates to a diversity of social and economic factors surrounding tool use and production. Here we explore the interaction of these factors on stone tool variability. As such, this study combines the pre-existing knowledge of stone tool production regarding, raw material properties, provenience and technology at Kanjera South with newly applied measures of stone tool utilization. This provides a novel means to understand Oldowan stone tool production from a multifaceted point of view in which the relationships between stone tool utilization, landscape-scale raw material constraints, and stone tool production strategies can be investigated.

```{r table}
rm <- read.csv("Data/Raw_mats.csv")
kable(rm,align = "c",caption = "A list of rock types included in this analysis")
```

   We characterize the technology of stone tools produced on both distant and local materials  at Kanjera South (Table 1) through the study of the core and complete flake assemblages (i.e. our analysis at this time does not incorporate angular fragments). Tool utilization for cores was measured in terms of reduction intensity. We estimate the proportion of mass lost prior to its discard following the methods outlined in [@douglassCoreUseLifeDistributions2017]. In addition, tool utilization is also examined by estimating the relative proportion of flakes in the Kanjera South assemblage that derive from early or later parts of the reduction sequence. We identify the sequence order that flakes were removed from their cores by using previously published models that use quantitative measures of flakes to predict their placement in a reduction sequence [@braunOldowanReductionSequences2008]. Flake sequence refers to the order that flake was removed from the core. Flake sequence data provides an independent measure of tool utilization that could be compared to the estimates of core reduction. Additionally, it also provides a way to examine the amount of flake production that occurred at Kanjera South. If cores were transported directly to Kanjera South and reduced, then the early flake sequences (i.e. the first flakes off the core) would be present in the assemblage. However, if early stage flakes are not present in the assemblage then it can be inferred that cores were utilized elsewhere prior to their transport to and discard at Kanjera South. Measures of core reduction intensity are analyzed along-side technological strategies (e.g. bidirectional) utilized in core reduction. Furthermore, core reduction and flake sequence values can also be examined according to raw material type. This provides a means to examine the impact of raw material properties and provenience on core reduction and technology. The integration of these multiple lines of evidence allows us to place stone tool variability at Kanjera south within the broader technological system in which it took place.


```{r include=FALSE}

cores$redux.strat <- "NA"
cores$redux.strat[cores$DeLaTorre.Typology == "Bidirectional Abrupt"] <- "Unifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Multifacial irregular"] <- "Multifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Multifacial irregularl "] <- "Multifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Unidirectional abrupt unipolar"] <- "Unifacial"
cores$redux.strat[cores$DeLaTorre.Typology ==  "Polyhedral "] <- "Multifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Unifacial centripetal exploitation "] <-  "Unifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Bifacial Partial Exploitation "] <-  "Bifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Unidirectional Abrupt bipolar "] <-  "Unifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "Unifacial simple partial exploitation "] <-  "Unifacial"
cores$redux.strat[cores$DeLaTorre.Typology == "bifacial centripetal"] <-  "Bifacial"


cores$DeLaTorre.Typology <- factor(as.character(cores$DeLaTorre.Typology), 
                                   levels = c("Unifacial simple partial exploitation ",
                                              "Unidirectional abrupt unipolar", 
                                              "Unidirectional Abrupt bipolar ",
                                              "Bidirectional Abrupt",
                                              "Bifacial Partial Exploitation ",
                                              "Unifacial centripetal exploitation ",
                                              "bifacial centripetal",
                                              "Polyhedral ",
                                              "Multifacial irregularl "))
```

## Characterizing Stone Tool Utilization.

   Tool utilization has been extensively researched in lithic analysis (see Andfresky 2009 for overview). Most studies of tool utilization focus on the degree of artifact reduction as a proxy for the use-life of an artifact (although these things are related but they are not the same; @shottExegesisCurationConcept1996). Reduction intensity can be calculated for a variety of artifact types [@dibbleMiddlePaleolithicScraper1995]. Currently several methods exist for estimating the extent of core reduction. The majority of these methods rely on establishing linear relationships between core attributes and the amount of mass lost during artifact production to predict mass removed from a core [@clarksonMeasuringCoreReduction2013; @douglassCoreUseLifeDistributions2017]. This study uses the methods outlined in Douglass et al. (2017) to estimate core reduction intensity as the proportion of mass lost.The predictive model developed by @douglassCoreUseLifeDistributions2017 incorporates a series of core attributes beyond simply surface area and total number of flake scars. This model considers, the number of flake scars, exploitation surfaces, the number of exploitation surface convergences, the proportion of cortex, and average platform angle to estimate core reduction intensity. 
   
   The _number of flake scars_ intuitively refers to the number of previous flake removals present on the core. The _number of exploitation surfaces_ refers to the number of areas of the core where flakes were removed along a similar axis. This variable is related to core rotation which is argued to increase as core reduction increases (e.g. Delagnes and Roche 2005). The _number of exploitation surface convergences_ documents the number of times different exploitation surfaces intersect with each other. Throughout reduction exploitation surfaces with different flaking axes tend to converge (Braun 2005, Douglass et al 2017). The proportion of cortex has an intuitive relationship with core reduction intensity. That is, the total proportion of the core that possesses cortical surface area will diminish during reduction. _Average platform angle_, measured in degrees, refers to the mean angle between striking surfaces. Various experimental replication studies show that, as this angle approaches 90°, it becomes increasingly difficult to detach a flake [@cotterellEssentialMechanicsConchoidal1985]. Therefore, as a core approaches exhaustion, the platform angles on the core are likely to approach 90° [@douglassCoreUseLifeDistributions2017]. Specifically, the method by @douglassCoreUseLifeDistributions2017 utilizes a generalized linear mixed model to examine the effects of predictor variables on core reduction intensity and ultimately predict the percentage of mass lost from a core. The added benefit of this model is that it is directly applicable to the materials analyzed in here, as it is was developed in validated on experimental material and a subset of archaeological material from Kanjera South [@braunOldowanReductionSequences2008]. These details are discussed in @douglassCoreUseLifeDistributions2017. 

## Flake Sequence Estimates

Flake sequence is a common analytical method in American archaeology [@andrefskyAnalysisStoneTool2009; @bamforthTechnologicalEfficiencyTool1986] but is also applied to Early Stone Age assemblages in various forms [@braunOldowanReductionSequences2008; @delatorreignacioEarlyStoneAge2011; @stoutTechnologicalVariationEarliest2010; @tothOldowanReassessedClose1985]. In the Early Stone Age, flake sequences are most commonly characterized using Toth types, which classifies flakes into six different stages depending on the presence of cortex on a flake’s platform and dorsal surface [@tothOldowanReassessedClose1985]. Here we follow @braunOldowanReductionSequences2008, using a multi-linear model to estimate flake sequence values. This methodology is specifically focused on understanding the approximate location of a flake within a reduction sequence. Unlike Toth’s flake types, which is focused on relative sequence information, the multi-linear model allows for an absolute placement of a flake within a reduction set (within a prescribed error). The multiple linear regression uses flake length, width, number of platform facets, number of flake scars, and the number of flake scar directions; specific details for each measurement are clearly outlined in @braunOldowanReductionSequences2008 (page 2156, figure 3). Before the sequence number can be estimated, the number of flake scars and amount of dorsal cortex must be divided by the log of the surface area of the flake [@braunOldowanReductionSequences2008]. These variables are then used by the predictive model to estimate the flake sequence number. Flake sequence estimates have a maximum error between +/- 8 sequences. However, an application of the method to refitting sequences from the Koobi Fora formation showed it always places flakes in their relative order[@braunOldowanReductionSequences2008]. Therefore, while information derived from individual flake sequence estimates may be coarse-grained, it remains useful for assemblage scale comparisons.

## Edge to mass ratios

Measurements of tool use are often associated with enumerating the details of the reduction of artifacts [@dibbleMiddlePaleolithicScraper1995; @erenKuhnGeometricIndex2009]. This is largely because of the reductive nature of stone technology (i.e. the process of producing stone tools always produces a smaller byproduct). However, some have also reviewed the general efficiency of this flaking process [@braunTechnologicalDevelopmentsOldowan2003; @brantinghamConstraintsLevalloisCore2001; @erenAreUpperPaleolithic2008; @mullerIdentifyingMajorTransitions2016]. Unfortunately, the study of efficiency in artifact production is not well standardized. There are some studies that investigate the overall efficiency of specific knapping strategies (Brantingham and Kuhn 2001; Eren et al. 2008). Others review efficiency as it pertains to individual flaking events within an overall technological framework [@braunTechnologicalDevelopmentsOldowan2003]. Identifying the mechanisms that result in highly efficient tool forms is not well understood [ @linUtilityEconomizationUnretouched2013; @dogandzicEdgeLengthSurface2015]. Yet it is clear that there is variation in how efficient some technologies are (even though the metric for identifying efficiency is debated; Eren et al. 2008). Assuming that most stone tools are produced to create sharp edges, one possible measure is estimating the amount of sharp edge produced per given unit of mass. Technologies that produced a higher amount of edge per volume of material can be considered more efficient (Braun and Harris 2003). Here we use a measure of edge that is based on tracing the edge of whole flakes [@braunTechnologicalDevelopmentsOldowan2003; @isaacKoobiForaResearch1997]. To calculate efficiency this edge estimate is divided by the logarithmic transformation of mass. The reason for this transformation is that mass increases in three dimensions (i.e. volumetrically) and the edge of a flake increases in two dimensions. The logarithmic transformation of mass prevents distortions of this ratio that are the result of general size parameters. For example, very small flakes have relatively high edge for a given amount of mass, but this is not always the most efficient way to produce the greatest amount of edge relative to volume  (e.g. see discussions on this topic in Kuhn 1991). Flakes that have high amounts of edged relative to their mass tend to be relatively thin flakes, and there is the possibility that the efficiency of these tools is limited by their capabilities to complete certain tasks (e.g. tacks that require intensive use of edges such as hide scraping may not be feasible with relatively thin flakes). Here we calculate the edge to mass ratio of flakes within raw material categories. We then aggregate this measure across raw materials types to assess the overall efficiency of a given raw material. These aggregate measures are likely more reflective of the generalized pattern of efficiency in tool production over time at the Kanjera South locality.

# Results

```{r read in results data, echo = FALSE, include=FALSE}
cores$RM <- toupper(cores$RM)
cores$rm.provenance <- "Distant"
cores$rm.provenance[cores$RM %in% c("HLI", "FNY", "HPH")] <- "Local"
cores <- cores[cores$RM != "NYC",]
flakes <- read.csv("Data/ksj_flakes_with_GLMM_Seq.csv")
```

```{r Kruskal Wallace cores,echo = FALSE, include=FALSE}

library("FSA")

kw.res <- kruskal.test(cores$fixedpredict, g = cores$RM)
core.kw.p <- kw.res$p.value 
dT.res <- dunnTest(cores$fixedpredict~ cores$RM)
```

```{r echo = FALSE, include = FALSE}
cores.table <- table(cores$rm.provenance,cores$DeLaTorre.Typology)
res <- fisher.test(cores.table, simulate.p.value = TRUE)

cores$redux.strat <- factor(x = cores$redux.strat, levels = c("Unifacial", "Bifacial", "Multifacial"))
cores$RM <- factor(cores$RM, levels = c("BBA", "BFE", "BQU", "NYR","OGR", "NYC", "HLI", "HPH", "FNY"))

no.flaws <- cores[!(cores$Reason.For.Discard %in% c("Flaw\nStep Fracture",
                                               "Step Fracture\nFlaw",
                                               "Angles \nFlaw",
                                               "Flaw",
                                               "Size\nFlaw",
                                               "Flaw\nAngles " ,
                                               "Flaw\nNone",
                                               "Angles \nSize\nFlaw",
                                               "Flaw\nSize")),]
```

## Core Utilization

   Core reduction intensity estimates reveal that there is a wide range of variation in the amount of mass that was removed from the cores at Kanjera South. Some cores were minimally utilized whereas others were reduced as much as 95% of their original mass. There is also a significant relationship between core reduction intensity and raw material type (Kruskal-Wallis, p = < .0001). This pattern appears to be driven by raw material provenance. Cores produced on raw material types that originate from more distant sources (BBa, BFe, BQu, NyR, and OGr) are on average more substantially reduced than those that occur locally (FNy, HPh, HLi) (Kruskal-Wallis, p < .001, Figure: \ref{core_redux_rm}).
   
```{r fig-4, echo = FALSE, warnings=FALSE, dpi= 1000,fig.height= 3, fig.width= 6, fig.cap= "The distribution of core reduction intensity values as predicted by the GLMM (Douglass et al 2017). The results show stark differences in the degree of reduction in materials originating from more distant sources than those that originate from local sources of stone. \\label{core_redux_rm}"}

a <- ggplot(cores, aes(x = RM, y = fixedpredict, fill = rm.provenance)) + 
   geom_boxplot(show.legend = FALSE) + 
   scale_fill_manual(values = c("#E69F00", "#56B4E9"))+
   scale_y_continuous(limits = c(0,1))+
   labs(x = "Raw Material Type", y = "Reduction intensity (% Mass Lost)") +
   theme(panel.background = element_rect(color = "black", fill = "white"))


b <- ggplot(cores, aes(x = rm.provenance, y = fixedpredict, fill = rm.provenance)) + 
   geom_boxplot(show.legend = FALSE) + 
   scale_fill_manual(values = c("#E69F00", "#56B4E9"))+
   scale_y_continuous(limits = c(0,1))+
   labs(x = "Raw Material Provenance", y = "Reduction intensity (% Mass Lost)")+
   theme(panel.background = element_rect(color = "black", fill = "white"))


### Curation Curves 
cores.rm <- ddply(cores, .(RM), summarize,
                            fixedpredict = unique(fixedpredict) ,
                            ecdf = ecdf(fixedpredict )(unique(fixedpredict)))




   
grid.arrange(a,b, ncol = 2)

```

   The application of the flake sequence model to the Kanjera South assemblage reveals a similar pattern to that found in the core reduction intensity analysis. Flake sequence values range from the first flakes off the core to the 30th flake in the sequence. As with core reduction intensity, raw material type has a significant influence on flake sequence values  (Kruskal-Wallis, p< .001). The largest differences are, again, those between rock types derived from more distant sources and those found locally. Flakes produced on rock types from more distant raw material sources are from later in the reduction sequence (figure \ref{flake_seq_rm}), while flakes from the locally found materials are from earlier stages of reduction \ref{flake_seq_rm}. Interestingly, there is a striking amount of homogeneity in the distribution of flake sequence values associated with exotic or distant raw materials. Aside from Bukoban Felsite (BFe) the inter-quartile range of flake sequence values are very similar from distant sources. Even though Bukoban Felsite has a wider range than the others, its median is nearly the same as the others. The flake sequence values associated with flakes from the local materials are also similar to each other but show slightly more variation. Homa Phonolite exhibits the widest range of flake sequence values. 
   
```{r fig-5, fig.height=3, fig.width=5, warning=FALSE, echo=FALSE, fig.cap= "The distribution of flake sequence values present within the Kanjera South flake assemblage. The primary differences in flake sequence values are between materials originating from more distant sources than those that originate from local sources of stone. \\label{flake_seq_rm}"}
### Doing Sequence the old way since there are currently problems with the GLMM ###

flakes$b.scar.dir <- flakes$Flake.scar.direction * .628


flakes$b.plat.facets <- flakes$Platform.facets * 1.008


flakes$b.scarXlogLXW <- flakes$fsbysa *.159
flakes$b.cortexXlogLXW <- flakes$dcbysa *.097
flakes$flake.stage <- flakes$b.scar.dir + flakes$b.plat.facets + flakes$b.scarXlogLXW + flakes$b.cortexXlogLXW
flakes$flake.stage <- flakes$flake.stage ^2


flakes$Raw.Material <- factor(toupper(flakes$Raw.Material), levels = c("BBA", "BFE", "BQU", "NYR","OGR", "NYC", "HLI", "HPH", "FNY"))
flakes$rm.provenance <- "Distant"
flakes$rm.provenance[flakes$Raw.Material %in% c("HLI", "FNY", "HPH")] <- "Local"

fs.res <- kruskal.test(flakes$flake.stage,flakes$Raw.Material)
fs.dunn.res <- dunnTest(flakes$flake.stage,flakes$Raw.Material)


a <- ggplot(flakes, aes(x = Raw.Material, y = flake.stage, fill = rm.provenance)) + 
   geom_boxplot()+
   scale_fill_manual(values = c("#E69F00", "#56B4E9"))+
   labs(x = "Raw Material Provenance", y = "Flake Sequence")+
   guides(fill=guide_legend(title="Raw Material Provenance")) +
   theme(panel.background = element_rect(color = "black", fill = "white"),
         legend.position = "right")
a
```

   As previously reported [@braunOldowanTechnologyRaw2009], the Kanjera core assemblage is comprised of a wide variety of technological types or core reduction strategies. The frequency of core reduction strategies present shows a significant relationship with the raw material type (Fishers exact test, p-value: < .001). In general, raw materials that derive from more distant sources, are represented by core reduction strategies that involve a greater number of core rotations or more complex rotations (i.e. centripetal flaking, figure: \ref{core.tech}). Though unifacial and unidirectional reduction strategies are present in small frequencies, there is a greater representation of centripetal, bifacial and multifacial exploitation strategies in materials from more distant origins (Figure: \ref{core.tech}). On the other hand, local materials such as the Fenetized Nyanzian (FNy) and Homa Phonolite (HPh) are represented by greater number of unifacial or uni-directional core reduction strategies (Figure: \ref{core.tech}). Homa Limestone, runs contrary to this general pattern. Although Homa Limestone is found in abundance locally, cores produced on this raw material type are often multi-facially reduced. However, as is addressed in the discussion this is likely related to the properties of the raw material itself [@braunOldowanTechnologyRaw2009]. When core reduction intensity is compared according by core reduction strategy an interesting pattern emerges. Unifacial and unipolar, core reduction strategies result in less reduction than strategies that require bifacial, multifacial or polyhedral strategies (Kruskal Wallis, P-value: < .001) (Figure: \ref{core.tech}). In other words, core reduction strategies that require fewer core rotations, such as unifacial and unidirectional strategies, are less reduced than those that involve more complex rotation strategies.
   
```{r fig-6,echo = FALSE, fig.width=6.5, fig.height= 4, fig.cap= "Left: The distribution of core reduction strategies by raw material type. With the exception of Homa Limestone, raw materials that derive from the Kisi highlights more greatly represented by complex core reduction strategies than those that can be found in the immediate vicinity of Kanjera South. Right: The distribution of reduction intensity values according to reduction strategy.  **USP**: Unifacial Simple Partial. **UAU**: Unidirectional abrupt unifacial. **UABI**: Unifacial abrupt bidirectional. **BA**: Bidirectional Abrupt. **BAP** Bifacial Partial. **UC**: Unifacial centripetal. **BC**: Bifacial Centripetal. **Poly**: Polyhedral. **MFI**: Mutifacial Irregular. The colors of the boxplots correspond with the representation of different reduction strategies in the left figure. \\label{core.tech}"}


cc <- scales::seq_gradient_pal("black", "light grey", "lab")(seq(0,1,length.out=9))

kruskal.res <- kruskal.test(cores$fixedpredict, cores$DeLaTorre.Typology)
dunn.res <- dunnTest(cores$fixedpredict, cores$DeLaTorre.Typology,method = "bonferroni")

redux.tech<-ggplot(cores, aes(x = DeLaTorre.Typology, y = fixedpredict, fill = DeLaTorre.Typology)) +
   geom_boxplot() +
   scale_fill_manual(values=cc) +
      labs(x = "Core Reduction Strategy", y = "Reduction intensity (% Mass Lost)")+
      guides(fill=guide_legend(title="Provenance")) +
   scale_x_discrete(labels= c("USP", "UAU", "UABI", "BA", "BAP", "UC", "BC","Poly", "MFI"))+
   theme(panel.background = element_rect(color = "black", fill = "white"),
         legend.position = "none",
         axis.text.x = element_text(vjust = 1, hjust = 1, angle = 70))
   

rm.tech <- ggplot(cores, aes(x =RM , fill = DeLaTorre.Typology)) +
   geom_bar(position = "fill") +
   scale_fill_manual(values=cc)+
   labs(x = "Raw Material", y = "Relative Frequency") + 
   theme(panel.background = element_rect(color = "black", fill = "white"),
         legend.position = "none")

grid.arrange(rm.tech, redux.tech,ncol=2)


redux.strat.table <- table(cores$RM,cores$DeLaTorre.Typology)

```

## Flake efficiency

Analysis of the relative proportion of edge to mass in flakes indicates significantly different technological strategies applied to the different raw materials from the Kanjera South assemblage. Although the mean values of the different raw materials are relatively similar, the overall distribution of values indicates that rock types from sources that are further away from Kanjera South (e.g. rhyolite, felsite, quartzite) are produced in a way that allows for much higher efficiency values than those seen in the rock types found close to Kanjera South. In particular, the relatively low values seen in the Fenetized Nyanzian and Homa limestone indicate that flakes made on these rock types were produced in a manner that does not increase the use life of these cores [@braunLandscapescaleVariationHominin2008; @shottExegesisCurationConcept1996; @shottModelingUseLifeDistributions2004]. In particular the differences between the rocks types from the Kisii highlands (Bukoban quartzite, felsite and basalt) show signficant differences from those rock types that can be found in the drainages near Kanjera (e.g. Homa Limestone and Fenetized Nyanzian, p<.01 for all pairwise comparisons between Homa Limestone and all Bukoban rock types; p<.001 for all comparisons between Fenetized Nyanzian and all other rock types Kruskal-Wallis Rank Sum test, chi2=312.70, df=5, pairwise comparisons between raw materials use Dunnn's Test with Benjamin-Hochberg correction for multiple comparisons). It should be noted that even though there are significant differences between the edge to mass ratios, the distributions show significant overlap (Figure \\ref{flake_efficiency}. This indicates that while it is physically possible to produce flakes with similar edge to mass ratios in each raw material type, hominins at Kanjera South did not implement this strategy as frequently on raw materials that were locally abundant. Hominins at Kanjera South consistently produced flakes with greater edge and less mass from rock types that came from more distant sources. We only include rock types that have greater than 50 flakes because the wide variance in values seen in this measure results in wilidly divergent values in small samples.

```{r fig-7, echo=FALSE,dpi= 1000,fig.height= 3, fig.width= 6, fig.cap= "Boxplots of the measures of flake efficiency. Y-axis represents perimeter of flakes divided by a logarithmically transformed mass value. \\label{flake_efficiency}"}

allflakes<-read.csv("~/Dropbox/JR_DRB/Kanjera_JHE/Kanjera_South/Kanjera_Lithics/PerMass_AllFlakes")

allflakes$RM <- factor(allflakes$RM, levels = c("BFe", "NYr", "BBA","BQu","HLi","FNy"), ordered = TRUE)

flakes$rm.provenance <- "Distant"
flakes$rm.provenance[flakes$Raw.Material %in% c("HLI", "FNY", "HPH")] <- "Local"

allflakes$rm.provenance<-"Distant"
allflakes$rm.provenance[allflakes$RM %in% c("HLi","FNy")]<-"Local"

Plot1<-ggplot(data=allflakes, aes(x=RM, y=PerMass, fill=rm.provenance))+
  geom_jitter()+
  scale_fill_manual(values = c("#E69F00", "#56B4E9"))+
  theme_classic()

Plot2<-Plot1+
  geom_violin()+
  ylab("Edge to Mass Ratio mm/log(grams)")+
  labs(fill = "Raw Material Provenance")+
  xlab("Raw Materials")


Plot2

#library(dunn.test)
#library(DescTools)

#Desc(PerMass~RM, allflakes, plotit=TRUE, main="Fleke Efficiency")

#TEST<-dunn.test(x=allflakes$PerMass,g=allflakes$RM, method="bh",)

```


# Discussion

## The Influence of Raw Material Properties

  The results of this study show an interaction between stone tool utilization, raw material type and core reduction strategies. The most striking distinction is the difference in the degree of utilization of materials from local and distant sources. That is, cores that are derived from the drainages of the more distant Kisi highlands are significantly more reduced than those from local sources. This is also reflected in the flake sequence data. Within the Kanjera South flake assemblage, those produced on exotic raw materials are more likely to be from later in the reduction sequence than flakes from locally sourced materials (Figure \ref{flake_seq_rm}). This difference can be explained by a few different factors. Firstly, raw material properties play a role in causing these differences. The local fenetized rocks (FNy) are often discarded before complete reduction because of the intersection of removals with step fractures or preexisting internal fracture planes present in the highly metasomatized rocks [@braunRawMaterialQuality2009]. The chalky nature and block-like geometry of Homa limestone (HLi) also limits the number of flakes that can be removed. In contrast, the majority of raw materials from more distant sources possess fewer flaws and fracture more predictably than those found locally [@braunRawMaterialQuality2009].
   
   However, differences in material properties do not explain all of the differences in reduction intensity between local and exotic materials. If all of the cores that were discarded due to presence of internal flaws are removed from the assemblage (figure \ref{core.tech}), the overall pattern remains. In addition, Homa phonolite, a local raw material, does not have the defects common in other local raw materials but it is still less reduced than the nonlocal raw materials. In contrast with the other exotic materials, the properties of the Oyugis granite (OGr) also constrains the removal of flakes. The particularly coarse grained nature of this rock type alters its fracture pattern. It is difficult to maintain angles of less than 90 degrees. It is likely that hominins could only exploit pre-existing angles on these granite blocks, thus limiting the degree that they could be reduced [@braunOldowanTechnologyRaw2009]. Despite the limitations imposed by its material properties, Oyugis granite is still more reduced than any of the local raw materials (figure \ref{core_redux_rm}). Previous research shows that the exotic raw materials that were transported to Kanjera South not only fracture more predictably but also produce flakes with more durable edges [@braunRawMaterialQuality2009; @delatorreOmoRevisitedEvaluating2004; @ludwigTechnologicalReassessmentEast1998]. The need to reduce relatively lower quality material may be related to the durability and the efficiency of the produced edges. In other words, flakes produced on local rock types have shorter use-lives because their edges dull more quickly. The consequence of this, is that for any given use, more flakes of a local raw material are required. In this sense, the high abundance of these materials and the relatively quick dulling of their cutting edges may explain the high frequency of local raw materials in the Kanjera South assemblage.  

## Implications for Land use

   The stark differences in the degree of core utilization according to raw material origin also reflects a dynamic hominin land-use system. Flake sequence values associated with exotic materials are often from later in the sequence than those associated with local raw materials. This pattern could indicate that cores from exotic sources were often utilized prior to their arrival and discard at Kanjera South. The presence of highly reduced exotic materials in contrast with lightly reduced local materials may imply that hominins that visited the Homa Peninsula may have discarded more utilized exotic materials as they collected materials that were proximal to Kanjera as part of a mobile toolkit. This pattern has been observed at the site of FxJj 50 from the Okote Mbr. of the Koobi Fora Fm. [@braunOldowanRawMaterial2009]. This may further suggest that stone tool acquisition and discard behavior may relate to the need to maximize the amount of utility hominins carried with them. **In this sense, hominins would have made decisions to discard raw material when opportunities to increase the amount useable volume was prssent.**  The core reduction pattern at Kanjera South is also consistent with expectations of raw material procurement and utilization when conditions are neutral. Under neutral conditions, raw materials from more distant sources will be, on average, more reduced, than those from local sources [@brantinghamNeutralModelStone2003; @popSimulatingLithicRaw2016]. The fact that cores from more distant sources are more reduced than those from local sources is consistent with this pattern. This "distance-decay" pattern has been observed in numerous archaeological studies (eg. @blumenschineEffectsDistanceStone2008; @closeDistanceDecayUneasy1999; @newmanEffectsDistanceLithic1994; @tothOldowanReassessedClose1985). In addition, neutral expectations state that the variance in tool reduction intensity will decrease with distance from a given source. In other words, the variance in the amount of stone tool reduction intensity should be substantially lower in assemblages of tools from distant sources. The converse is also predicted by the distance-decay model. Variance in core reduction will be much greater in local sources. 
   
   This prediction of the distance-decay model is not supported by the Kanjera South data. In contrast, while exotic materials are reduced more more substantially than local materials, the interquartile ranges of flake sequence and core reduction measures of assemblages from distant sources are as wide (or wider) than those associated with local sources. This argues against a null hypothesis that the patterns seen in the lithic assemblage at Kanjera South formed under neutral conditions (i.e. no intentional attempt to increase the use life of artifacts from distant sources). It has been hypothesized that deviations from the neutral model of this nature may arise due to increasingly linear movements toward specific locations [@brantinghamNeutralModelStone2003; @braunLandscapescaleVariationHominin2008]. Moreover, subsequent work modeling the influence of directed movement towards attractors has shown that while a distance-decay pattern remains visible, tools from earlier stages of reduction will be over-represented (i.e. greater variance in reduction) [@reevesDigitalStoneAge2019]. Thus, the greater than expected range in  variance in the reduction intensity of distantly sourced cores may suggest that hominins directed their movement to Kanjera South. 
   
   The concept that hominins directed their movement toward Kanjera South, is supported by other archaeological and paleoecological evidence. Numerous taphonomic studies of the faunal assemblage from Kanjera South have verified that hominins efficiently exploited small bovids and may have processed larger carcasses that were scavenged from carnivores. [@ferraroEarliestArchaeologicalEvidence2013; @oliverBovidMortalityPatterns2019]. Use-wear studies that demonstrate that hominins carried out a variety of resource processing activities with the stone artifacts that were produced at Kanjera South, including butchery and the processing of a variety of plants, including underground storage organs (USOs)  [@lemoriniOldStonesSong2014; @lemoriniOldStonesSong2019]. These studies attest to the resource rich environment in which Kanjera South was situated. These resources ultimately made the Kanjera South locale and the Homa Pennisula an attractive place for hominins, most likely for centuries given that the finds were made through three meters of sequence (upper KS-1 through KS-3).  Thus, the patterns evinced by this study are the result of the repeated visitation by hominins over generations to Kanjera South, and these patterns likely reflect the broader foraging ecology of Oldowan Tool makers.
   
## The interacting effect of land-use and material properties on lithic technology at Kanjera South.
  
   The results of this study may also suggest that tool utilization, raw material properties and foraging ecology may interact in complex ways that structure the patterns of artifact variation we see at Oldowan sites. Kanjera South artifacts include a wide range of reduction strategies [@braunOldowanTechnologyRaw2009; @plummerEnvironmentalContextOldowan2009; @plummerLatePlioceneOldowan2001; @plummerOldowanHomininBehavior2016]. These strategies have a similar appearance to the descriptions of tool forms that have been described in other Oldowan contexts [@delatorreTechnologicalStrategiesLower2005]. Interestingly the different reduction strategies present at Kanjera South seem to be related to reduction intensity. Cores that are described as having bifacial, multifacial and/or centripetal patterns of removals exhibit high levels of reduction.  Few of the unifacial cores and partial bifacial cores show this increased level of reduction intensity (Figure \ref{core.tech}). It could be argued that this pattern is a consequence of the predictive model used to estimate core reduction intensity. The reduction intensity model predicts higher reduction intensity as core rotation increases. This is a pattern that has been identified in some experimental models of Oldowan core reduction [@douglassCoreUseLifeDistributions2017]. However, this is not necessarily the case. Despite the fact that cores produced on Oyugis Granite and Homa Limestone are often irregular and multi-facial, the predictive model still discriminates between different levels of reduction in these cores. Complimentary support can be found in the flakes made from these raw materials. Flakes made from exotic cores exhibit flake sequence values that suggest they were produced from later in the flaking sequence. This independent line of evidence suggests that the predictive model of core reduction is not over estimated due to rotation. Rather, this model of reduction intensity is accurately identifying estimates of the length of core use life in these exotic cores. Therefore it is possible to examine the representation of various core reduction strategies in terms of raw material properties and land-use. 
   
   The representation of reduction strategies in each raw materials type also follows the interaction between raw material properties and broader land-use that is described above. Raw material properties have been shown to influence reduction intensity decisions previously [@braunOldowanTechnologyRaw2009]. The frequent application of unifacial reduction strategies to the locally obtained FNy may result from the limitations imposed by the internal flaws often present in the rock. In other words, hominins may have only been able to exploit a single surface on these metasomatized rocks before the knapping process  encountered an internal flaw. Rather than attempt to overcome the flaw by rotating the core, the high abundance of FNy at KJS may have made it easier to simply discard the core and acquire a new nodule. High discard rates in contexts of abundant raw material has been observed in numerous archaeological and ethnographic contexts [@andrefskyRawMaterialAvailabilityOrganization2008]<!-- Fact check this -->. Moreover, the technological analysis of the cores suggests that the high representation of irregular multi-facial cores produced in Homa Limestone (HLi) and Oyugis Granite (OGr) are the result of continuous core rotation to exploit naturally occurring angles [@braunOldowanTechnologyRaw2009]. The exploitation of naturally occurring angles in rocks is a fairly common feature of Oldowan hominins. 
[@delagnesLatePlioceneHominid2005]. Yet these mechanical issues do not explain all of the patterns seen at Kanjera South. Homa phonolite is one of the most homogeneous raw materials found at Kanjera South. Hominins at Kanjera South frequently implemented bifacial reduction strategy in these cores. It was possible to implement longer reduction chains in this material than was possible with the other local materials. Yet the these cores are never reduced to the same degree as that seen in the exotic raw materials. 
   
   The cores produced on raw materials from the area surrounding Kanjera South (i.e. local materials) exhibit an even representation of various core reduction strategies. The same pattern is not seen in rocks transported over large distances. Exotic raw materials show a strong bias toward more reduced, complete bifacial and multifacial reduction strategies (Figure \ref{core.tech}). While the HPh shows a high frequency of bifacial reduction, this reduction is only partial (Figure \ref{core.tech}). Moreover, cores with multi-facial or centripetal flaking strategies are seldom produced on Homa Phonolite (Figure \ref{core.tech}). This contrasts sharply with the prevalence of more intense core reduction strategies involving greater degrees of core rotation in rock types from further sources (e.g. Bukoban quartzite). This suggests that raw material properties is not the only factor influencing core reduction strategies. 
   
   The results of the edge-to-mass ratio data suggest that hominins applied a strategy that more efficiently produced flakes on rock types transported from greater distances from Kanjera South. In addition, previous research shows that exotic raw materials also produce more durable cutting edges [@braunRawMaterialQuality2009]. These two factors may have incentivized the retention of exotic raw materials within the hominin tool-kit. The stock of these "high quality" raw materials would have become increasingly more expensive (i.e. in time and energy) to rejuvenate as hominins moved away from the Kisi highlands [@braunOldowanBehaviorRaw2008]. In this light, the high frequency of the bifacial and multifacial reduction strategies may have arisen from a need to maximize the amount utility that could be extracted from these cores. The exploitation of multiple flake removal surfaces on a core allows a core to remain active in a toolkit for a longer period of time. In later times hominins more frequently retouched formalized tools when the availability of raw materials becomes scarce [@clarkLithicsLandscapesLongueduree2017]. In a similar sense, the application of complete bifacial and multifacial strategies may have allowed hominins to utilize these high quality exotic materials longer to optimize resource acquisition at Kanjera South. 
   
   It could be argued that this pattern reflects the high level planning and foresight required to utilize and transport tools within this dynamic system. However, simpler explanations should also be considered. Modeling work has shown that tools will inevitably become more utilized as they are transported from the source, even under neutral or random conditions [@brantinghamNeutralModelStone2003; @popSimulatingLithicRaw2016; @reevesDigitalStoneAge2019]. Experimental research has also shown that the wide range of formal types defined by Leakey (1971) could arise through a least effort flaking strategy [@tothOldowanReassessedClose1985; @tothStoneTechnologiesEarly1982]. More recently, @mooreExperimentalInsightsCognitive2016 showed cores will eventually become bifacial or even multi-facial when the flaking procedure is systematically randomized. Thus, it may be possible to produce the differences in reduction intensity and reduction strategies observed between local and exotic materials by simply combining the influence of simple flaking procedures, stone tool transport, and raw material selection. Given the relationship between core reduction patterns and stone tool reduction intensity, the data from Kanjera South may suggest that different reduction strategies reflect different points on a reduction continuum rather than intentionally imposed strategies or socially mediated patterns of production. A similar notion has been proposed from LCT variability [@iovitaHandaxeReloadedMorphometric2011]. Oldowan researchers must consider the possibility that, in some cases, Oldowan core variability arises due to an interaction of simple flaking rules with broader ecological parameters. In this sense, some Oldowan core variation may be an emergent phenomenon, rather than a direct reflection of the intentions or information that was transmitted to the knapper.
   
```{r fig-8, eval=FALSE, fig.cap=, fig.width=6.5, include=FALSE}

a <- ggplot(cores, aes(x = DeLaTorre.Typology, fill = rm.provenance)) + 
   geom_bar(aes(y = (..count..)),position = "dodge") + 
   scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
   guides(fill = guide_legend(title = NULL)) + 
   theme(panel.background = element_rect(color = "black", fill = "white"),
         axis.text.x = element_blank(),
         axis.title = element_blank(),
         legend.position = c(.085,.75),
         legend.background = element_blank(),
         legend.title = element_text(size = 3))

txt <- "Relative frequency of reduction strategies of\ncores not discarded due to flaws in the rock"
b <-  ggplot(no.flaws, aes(x = DeLaTorre.Typology, fill = rm.provenance)) +
   geom_bar(aes(y = (..count..)),position = "dodge") + 
   scale_fill_manual(values = c("#E69F00", "#56B4E9")) +
   annotate("text", x = 2.45, y = 15, label = txt, size = 3)+
   theme(panel.background = element_rect(color = "black", fill = "white"),
         axis.text.x = element_blank(),
         axis.title = element_blank(),
         legend.position = "none")

grid.arrange(a,b,
             top = textGrob("Relative frequency of reduction strategies according to by raw material provenance"),
             left = textGrob("Relative frequency", rot = 90))
```

# Implications for the Oldowan as a whole. 

  The site of Kanjera South is unique in relation to Oldowan sites of a similar age or older. Many of the sites 2 million years old or older are situated in close proximity to raw materials sources and are comprised of a singular reduction strategy [@rocheOriginsAdaptationsEarly2009; @plummerArchaeologicalSitesMa2018]. In contrast, the assemblage at Kanjera South shows a substantial representation of exotic raw materials and a diversity of different core reduction strategies. The results from Kanjera South elucidate how the interaction of raw material properties, foraging ecology, and landscape scale constraints on raw material availability influence technological variability in the Oldowan. While many sites show strong selection for specific materials, the frequent use of unifacial reduction strategies at sites such as Lokalelei 2C, East Gona, Hadar, Omo, Ledi Geraru may relate to the overall abundance of knappable material that is immediately available at these site [@braunEarliestKnownOldowan2019; @kimbelLatePlioceneHomo1996; @rocheEarlyHominidStone1999; @stoutRawMaterialSelectivity2005]. In other words, there maybe little incentive to exhaustively reduce a core when material is so abundant [@clarkLithicsLandscapesLongueduree2017].

  Kanjera South is also the only site from this time frame that is situated in an open grassland [@plummerOldestEvidenceToolmaking2009]. Humans tend to travel farther and more frequently in open arid environments than those that live in more productive environments [@burnsideHumanMacroecologyLinking2012]. Moreover savanna adapted Chimpanzees from Fongoli also possess a larger home range and practice fission-fusion less frequently [@pruetzChimpanzeePanTroglodytes2009]. If this was also the case for the hominins at Kanjera South, these factors may have required new technological strategies in order to accommodate expanded home range sizes. On a broader scale the technological diversity represented at Kanjera South may reflect adaptations to the specific contextual conditions at this locale. The stimulus for increasing the use lives of cores (through reduction intensity and efficient use of raw material) at Kanjera South may have been foraging activities that quickly dulled artifact edges (e.g., butchery and USO processing) [@braunOldowanBehaviorRaw2008; @lemoriniOldStonesSong2014] that were more frequently carried out in open habitats  [@ferraroEarliestArchaeologicalEvidence2013; @lemoriniOldStonesSong2019; @plummerEnvironmentalContextOldowan2009]. 
  
  Finally while the preceding analysis emphasizes the role of the broader environment and land-use on technological variability, ecology is not the sole driver of Oldowan technical variation. The inter-quartile ranges in figure \ref{core.tech} show a substantial amount of overlap between the reduction intensity and core reduction strategies. This suggests, that not all variation can be explained by environmental parameters such as raw material avalaibility and material properties. Moreover, inter-site differences between West Turkana sites are not easily explained factors such as raw material availability alone [@rocheNaiyenaEngolWest2018]. This unexplained variation may be the result of socio-cultural dynamics that may have maintained information regarding the stone tool production process between groups. However, the fidelity and the mechanisms that underlie the maintenance of this information remain an open debate [@hoversInventionReinventionInnovation2012; @morganExperimentalEvidenceCoevolution2015; @stoutArchaeologyOriginsHuman2019; @tennieEarlyStoneTools2017; @tennieIslandTestCumulative2016]. However, the results of this analysis show that it may be counterproductive for research on Oldowan technology to focus on specific aspects of stone tool technology (Figure \ref{variation}). Rather studies should examine the Oldowan within the broader socio-ecological system. It may be the interaction between function, ecology and sociality that provide information on the cultural capacity of Oldowan tool makers. 
   
# Conclusion  

Despite the superficial simplicity of the Oldowan, its variability reflects a complex interconnection of ecological, behavioral and social factors. These factors are inextricably linked even though researchers often study them components in isolation. The lithic assemblage from Kanjera South provides an excellent opportunity to examine the interacting effects of raw material properties, provenance and availability of rock types and presumably social variation in the form of Oldowan artifacts. This is especially important given the context of other information we have from Kanjera that indicates that hominins butchered multiple small animals at this site and engaged in other subsistence strategies. Combination of quantitive measures of stone tool reduction with qualitative characterizations of lithic technology (e.g @braunOldowanTechnologyRaw2009; @plummerOldowanHomininBehavior2016) provides new insights into the ecological factors that influence oldowan technology, and hominin behavior. 
   
   The fact that exotic materials are much more substantially reduced that from local sources reflects the differences in the quality of the raw materials available. The higher quality of the exotic materials would have incetivized the transport of these materials over longer distances. In this light, these differences in reduction also highlight that Oldowan tools were part of a mobile tool kit that reflects a broader land-use strategy. The marked differences in reduction intensity in combination with the paucity of early sequence flakes also suggests that exotic materials were often utilized prior to their arrival at Kanjera south. Although exotic materials are more reduced than local marerials the variance in the amount of stone tool reduction exhibited in the Kanjera South assemblage does not adhere to neutral expectations. This result suggests that the lithic assemblage at Kanjera South reflects a structured land-use pattern, that may indicate that hominins directed their movement, at least on occassion, to Kanjera South.

   This pattern also appears to have an influence on the technological strategies employed by Oldowan tool makers at Kanjera South. The relationship between core reduction strategies and reduction intensity indicates that raw material quality and provenance have a strong influence on the technological variation observed within a lithic assemblage. While these results show that ecological parameters have a strong effect on stone tool variation, a substantial amount of variation remains unexplained by ecology alone. Overall, these results suggest that future studies must utilize a more integrated approach to understand the behavioral significance of the Oldowan.

# Acknowledgements

The authors thank Claudio Tennie for reading an earlier version of the paper. Reeves also thanks the STONECULT Project for supporting his research. The project STONECULT has received funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No. 714658).

# References